Declarative Methods Term Project: \\ JumanG

### 2013-05-04 doc/paper.tex
final version for submission

\subsection{Radial Solver}
Radial displays have been used to display trees in a way different from a typical top-down approach.
Introduced in 1992 by Eades \cite{freeTrees}, this layout places the branches circularly around the root outwards, drawing 
nodes at different radii from the center. We implemented our own Radial solver that uses similar techniques to the original
method. We wrote our solver to evenly distribute nodes around the root at each level, rather than the typical style of weighting 
the nodes based on the subnodes. We also chose to display nodes at their closest level to root. 
This allows the solver to draw non-tree graphs fairly well. In some instances, we can even beat Graphviz's twopi instantiation, detailed below in our experimental comparison section. %TODO wording

@@

Here are some results based on our radial solver:

\begin{figure}
\caption{Radial Solver on a graph with high branching factor}
\begin{tikzpicture}[scale=0.40, transform shape]
	\tikzstyle{every node} = [circle, fill=gray!30, minimum size = 1.4cm]
	\node (24) at (3.52, 8.28) {24};
	\node (25) at (1.25, 8.91) {25};
	\node (26) at (-8.34, -3.37) {26};
	\node (27) at (-7.55, -4.90) {27};
	\node (20) at (1.55, -5.80) {20};
	\node (21) at (3.86, -4.60) {21};
	\node (22) at (5.20, -3.00) {22};
	\node (23) at (5.91, -1.04) {23};
	\node (28) at (-6.89, -5.79) {28};
	\node (29) at (7.19, -5.42) {29};
	\node (1) at (0.00, 0.00) {1};
	\node (3) at (0.00, 3.00) {3};
	\node (2) at (2.60, 1.50) {2};
	\node (5) at (-2.60, -1.50) {5};
	\node (4) at (-2.60, 1.50) {4};
	\node (7) at (2.60, -1.50) {7};
	\node (6) at (0.00, -3.00) {6};
	\node (9) at (5.56, 2.25) {9};
	\node (8) at (5.96, 0.73) {8};
	\node (11) at (3.69, 4.73) {11};
	\node (10) at (4.79, 3.61) {10};
	\node (13) at (-1.55, 5.80) {13};
	\node (12) at (1.55, 5.80) {12};
	\node (15) at (-5.96, -0.73) {15};
	\node (14) at (-5.20, 3.00) {14};
	\node (17) at (-4.79, -3.61) {17};
	\node (16) at (-5.56, -2.25) {16};
	\node (19) at (-1.55, -5.80) {19};
	\node (32) at (8.86, -1.56) {32};
	\node (31) at (8.16, -3.80) {31};
	\node (30) at (7.71, -4.64) {30};
	\node (18) at (-3.69, -4.73) {18};
	\draw [->] (22)--(29);
	\draw [->] (22)--(30);
	\draw [->] (22)--(31);
	\draw [->] (23)--(32);
	\draw [->] (1)--(2);
	\draw [->] (1)--(3);
	\draw [->] (1)--(4);
	\draw [->] (1)--(5);
	\draw [->] (1)--(6);
	\draw [->] (1)--(7);
	\draw [->] (3)--(12);
	\draw [->] (3)--(13);
	\draw [->] (2)--(8);
	\draw [->] (2)--(9);
	\draw [->] (2)--(10);
	\draw [->] (2)--(11);
	\draw [->] (5)--(15);
	\draw [->] (5)--(16);
	\draw [->] (5)--(17);
	\draw [->] (5)--(18);
	\draw [->] (4)--(14);
	\draw [->] (7)--(21);
	\draw [->] (7)--(22);
	\draw [->] (7)--(23);
	\draw [->] (6)--(19);
	\draw [->] (6)--(20);
	\draw [->] (12)--(24);
	\draw [->] (12)--(25);
	\draw [->] (17)--(27);
	\draw [->] (17)--(28);
	\draw [->] (16)--(26);
\end{tikzpicture}
\end{figure}

@@

The Springs algorithm, as outlined by Kawai and Kamada\cite{springs}, computes a layout for a graph by running a physics simulation on the graph. All of the nodes are set to repel each other, while the edges are set as ``springs" to hold the nodes together. Our solver scales the repelling forces of the nodes based on how densely connected the graphs are, in order to prevent nodes being spaced too close or far apart in the end.

@@

\subsubsection{Springs solver's Initialization}
The Springs solver also requires the nodes to have an initial configuration. Wilkinson \cite{grammarGraphics}
%TODO style?
suggests two possible approaches. The first being random placement, where all of the nodes are randomized, and the second being based on geodesic distance. We have implemented both options, the second option using the Radial solver as an alternative to this geodesic distance initialization.

@@

We measured connectivity as:
$$connectivity = \frac{|nodes|}{|edges|}$$
And based on some empirical testing we determined to use randomized initialization on graphs with connectivity higher than 3.

@@

\subsection{Layered Solver}
Layered graph, or Sugiyama-style drawing is an approach to arranging directed acyclic graphs\cite{sugiyama}. This drawing approach relies on assigning the nodes to a ``layer" in the graph and then assigning descending y-coordinates for each successive layer from the root. Then, for each layer, the nodes are rearranged to minimize the edge crossings in between layers.

@@

After determining all of the layers, the nodes in each layer are rearranged in an attempt to minimize edge crossings. Determining an ordering that minimizes edge crossings is NP-Complete\cite{EdgeCrossings}, and so we rely on heuristics to accomplish this task.

@@

\subsubsection{Edge Crossing Minimization Heuristic}
The heuristic our Layerd Solver used to order nodes and minimize edge crossings between layers relies on a calculation analogous to ``center of mass" using the edges between the surrounding layers. Each node is assigned an averaged index interpolating between the indexes of all the adjacent nodes above and below the layer, and then the nodes are sorted based on these indexes.

@@

\bibliographystyle{plain}
\bibliography{citations}

>>>>

\begin{figure}
\caption{Radial Solver on a graph with high branching factor}
\label{sprawl}
\begin{tikzpicture}[scale=0.40, transform shape]
	\tikzstyle{every node} = [circle, fill=gray!30, minimum size = 1.4cm]
	\node (24) at (3.52, 8.28) {24};
	\node (25) at (1.25, 8.91) {25};
	\node (26) at (-8.34, -3.37) {26};
	\node (27) at (-7.55, -4.90) {27};
	\node (20) at (1.55, -5.80) {20};
	\node (21) at (3.86, -4.60) {21};
	\node (22) at (5.20, -3.00) {22};
	\node (23) at (5.91, -1.04) {23};
	\node (28) at (-6.89, -5.79) {28};
	\node (29) at (7.19, -5.42) {29};
	\node (1) at (0.00, 0.00) {1};
	\node (3) at (0.00, 3.00) {3};
	\node (2) at (2.60, 1.50) {2};
	\node (5) at (-2.60, -1.50) {5};
	\node (4) at (-2.60, 1.50) {4};
	\node (7) at (2.60, -1.50) {7};
	\node (6) at (0.00, -3.00) {6};
	\node (9) at (5.56, 2.25) {9};
	\node (8) at (5.96, 0.73) {8};
	\node (11) at (3.69, 4.73) {11};
	\node (10) at (4.79, 3.61) {10};
	\node (13) at (-1.55, 5.80) {13};
	\node (12) at (1.55, 5.80) {12};
	\node (15) at (-5.96, -0.73) {15};
	\node (14) at (-5.20, 3.00) {14};
	\node (17) at (-4.79, -3.61) {17};
	\node (16) at (-5.56, -2.25) {16};
	\node (19) at (-1.55, -5.80) {19};
	\node (32) at (8.86, -1.56) {32};
	\node (31) at (8.16, -3.80) {31};
	\node (30) at (7.71, -4.64) {30};
	\node (18) at (-3.69, -4.73) {18};
	\draw [->] (22)--(29);
	\draw [->] (22)--(30);
	\draw [->] (22)--(31);
	\draw [->] (23)--(32);
	\draw [->] (1)--(2);
	\draw [->] (1)--(3);
	\draw [->] (1)--(4);
	\draw [->] (1)--(5);
	\draw [->] (1)--(6);
	\draw [->] (1)--(7);
	\draw [->] (3)--(12);
	\draw [->] (3)--(13);
	\draw [->] (2)--(8);
	\draw [->] (2)--(9);
	\draw [->] (2)--(10);
	\draw [->] (2)--(11);
	\draw [->] (5)--(15);
	\draw [->] (5)--(16);
	\draw [->] (5)--(17);
	\draw [->] (5)--(18);
	\draw [->] (4)--(14);
	\draw [->] (7)--(21);
	\draw [->] (7)--(22);
	\draw [->] (7)--(23);
	\draw [->] (6)--(19);
	\draw [->] (6)--(20);
	\draw [->] (12)--(24);
	\draw [->] (12)--(25);
	\draw [->] (17)--(27);
	\draw [->] (17)--(28);
	\draw [->] (16)--(26);
\end{tikzpicture}
\end{figure}

\subsection{Radial Solver}

Radial displays have been used to display trees in a way different from a typical top-down approach.
Introduced in 1991 by Eades \cite{Eades:91} %~\cite{freeTrees}, 
this layout places the branches circularly around the root outwards, drawing 
nodes at different radii from the center. We implemented our own Radial solver that uses similar techniques to the original
method. We wrote our solver to evenly distribute nodes around the root at each level, rather than the typical style of weighting 
the nodes based on the subnodes. We also chose to display nodes at their closest level to root. 
This allows the solver to draw non-tree graphs fairly well. In some instances, we can even beat Graphviz's twopi instantiation, detailed below in our experimental comparison section. %TODO wording

@@

You can see a graph drawn using our radial solver in Figure~\ref{sprawl}.

@@

The Springs algorithm, as outlined by Kawai and Kamada \cite{Kamada:89}, computes a layout for a graph by running a physics simulation on the graph. All of the nodes are set to repel each other, while the edges are set as ``springs" to hold the nodes together. Our solver scales the repelling forces of the nodes based on how densely connected the graphs are, in order to prevent nodes being spaced too close or far apart in the end.

@@

\subsubsection{Springs solver's Initialization}
The Springs solver also requires the nodes to have an initial configuration. Wilkinson suggests two possible approaches \cite{Wilkinson:05}. The first being random placement, where all of the nodes are randomized, and the second being based on geodesic distance. We have implemented both options, the second option using the Radial solver as an alternative to this geodesic distance initialization.

@@

We measured connectivity as:
$$connectivity = \frac{|nodes|}{|edges|}$$
Based on empirical testing we determined to use randomized initialization on graphs with connectivity higher than 3.

@@

\subsection{Layered Solver}
Layered graph, or Sugiyama-style drawing is an approach to arranging directed acyclic graphs \cite{Sugiyama:81}. This drawing approach relies on assigning the nodes to a ``layer" in the graph and then assigning descending y-coordinates for each successive layer from the root. Then, for each layer, the nodes are rearranged to minimize the edge crossings in between layers.

@@

After determining all of the layers, the nodes in each layer are rearranged in an attempt to minimize edge crossings. Determining an ordering that minimizes edge crossings is NP-Complete \cite{Eades:94}, and so we rely on heuristics to accomplish this task.

@@

\subsubsection{Edge Crossing Minimization Heuristic}
The heuristic our Layered Solver used to order nodes and minimize edge crossings between layers relies on a calculation analogous to ``center of mass" using the edges between the surrounding layers. Each node is assigned an averaged index interpolating between the indexes of all the adjacent nodes above and below the layer, and then the nodes are sorted based on these indexes.

@@

%\bibliographystyle{plain}
%\bibliography{citations}
\begin{thebibliography}{}

%springs
\bibitem[\protect\citename{Kamada and Kawai}1989]{Kamada:89}
Tomihisa Kamada and Satoru Kawai.
\newblock 1989.
\newblock {\em An Algorithm for Drawing General Undirected Graphs}, volume~31.
\newblock Information Processing Letters.

%grammerGraphics
\bibitem[\protect\citename{Wilkinson}2005]{Wilkinson:05}
Leland Wilkinson
\newblock 2005.
\newblock {\em The Grammar of Graphics}.
\newblock Springer Science+Business Media, Inc.

%radial
\bibitem[\protect\citename{Eades}1991]{Eades:91}
Peter Eades
\newblock 1991.
\newblock {\em Drawing Free Trees}.
\newblock International Institute for Advanced Study of Social Information Science, Fujitsu Limited.

% edgecrossings is NP
\bibitem[\protect\citename{Eades}1994]{Eades:94}
Eades, Peter and Wormald, Nicholas C.
\newblock 1994.
\newblock {\em Edge crossings in drawings of bipartite graphs}, volume~11.
\newblock Springer-Verlag.

%layered
\bibitem[\protect\citename{Sugiyama}1981]{Sugiyama:81}
Kozo Sugiyama
\newblock 1981.
\newblock {\em Methods for Visual Understanding of Hierarchical System Structures}, volume~11.
\newblock IEEE Transactions on Systems, Man, and Cybernetics.
